---
title: "tbd"
author: "Philipp B."
date: "13 6 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

knitr::opts_chunk$set(fig.path = "figure/fig-")

Sys.setlocale(category = "LC_ALL", locale = "english")
```

# Synopsis
1. Across the United States, which types of events (as indicated in the <p style="color:red">EVTYPE</p> variable) are most harmful with respect to population health?
1. Across the United States, which types of events have the greatest economic consequences?

# Data Processing

packages
```{r, results='hide', message=FALSE}
library(lubridate)
library(dplyr)
library(ggplot2)
```

```{r, cache=TRUE}
if (!file.exists("repdata_data_StormData.csv.bz2")) {
    download.file(url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                  destfile = "repdata_data_StormData.csv.bz2")
}
noaa <- as_tibble(read.csv("repdata_data_StormData.csv.bz2"))
```

To do:
- convert BGN_DATE and BGN_TIME, END_DATE and END_TIME to date/time using time zones
- check documentation on BGN_TIME
- county, state, end_county and evtype as factor

test_dt2 <- mdy_hm(paste(gsub(" 0:00:00", "", noaa$BGN_DATE), noaa$BGN_TIME, sep = " "))
test_dt3 <- mdy_hms(noaa$BGN_DATE)
test_dt4 <- mdy_hms(paste(sub(" 0:00:00", "", noaa$BGN_DATE), na.omit(noaa$BGN_TIME), sep = " "))

Test:
noaa[match(unique(noaa$TIME_ZONE), noaa$TIME_ZONE), c("BGN_DATE", "BGN_TIME", "TIME_ZONE", "BGN_DT")]

```{r parsing date}
noaa <- noaa %>% 
    mutate(BGN_DT = if_else(grepl("(PM|AM)", BGN_TIME, ignore.case = T), 
                           mdy_hms(paste(sub(" 0:00:00", "", BGN_DATE), noaa$BGN_TIME, sep = " "), quiet = T),
                           mdy_hm(paste(sub(" 0:00:00", "", BGN_DATE), noaa$BGN_TIME, sep = " "), quiet = T))) %>% rowwise %>%
    mutate(BGN_DTTZ = if_else(grepl("(PM|AM)", BGN_TIME, ignore.case = T), 
                           mdy_hms(paste(sub(" 0:00:00", "", BGN_DATE), noaa$BGN_TIME, sep = " "), tz = TIME_ZONE, quiet = T),
                           mdy_hm(paste(sub(" 0:00:00", "", BGN_DATE), noaa$BGN_TIME, sep = " "), tz = TIME_ZONE, quiet = T)))
```

# Results